\chapter{Introduction}
Cyber-infrastructural systems are getting more integrated with our day-to-day life and have more impact on the society, thus, there is a need for these systems to be designed in a way that are more aligned with norms of society (or policies, rules, regulations). This means norms and policies should be considered as part of system development and maintenance cycles by the designers and developers. On the other hand, systems affect the society and by extension they are also affecting the norms, so they should be already considered when policy-makers are analyzing existing norms or implementing new ones. Taking modelling and model execution as a principal approach to designing both systems and norms, intuitively, requires the designers to be able to create models of these concepts, which include executable models the system, relevant norms, and, the social setting and its actors. The expressivity and flexibility required to encompass these aspects is not trivial. It is no longer just the model of a system, a social setting, and set of norms, but, it is a system that is utilized by social actors and governed by norms, a social setting that utilize and monitor a system, and decide upon, modify, analyze and enforce the norms, and, norms that regulate the system and the social setting and have impacts on both.


The purpose of this thesis is to closely study the interactions between these concepts, search for fundamental gaps in current methodologies utilized by system designers and policy makers, with the goal of finding approaches for creating more expressive and tangible models that are also flexible enough for specifying different scenarios and use-cases in different domain to assist in both system design and policy-making. To propose realistic methods, aside from the theoretical analysis, proof-of-concept tools have also been developed or adopted at each step and are presented.


\paragraph{Social Norms and Behavioral Specifications}
A central theme in this thesis are norms and their relation with individual behavioral specifications. The definition of norms for specific use-cases and domains is specified in corresponding chapters, but, it is valuable to set the tone on the complexities of this concept. The daily social life of a human is filled with objectively complex interactions, but somehow we are able to navigate these dynamic and nondeterministic interactions with sufficiently low allocation of processing resources. A concept that aids us in doing so is norms, by defining what should and should not be the case and what is expected to happen or not to happen in certain contexts and conditions, norms drive the actions in our fairly complex interactions. 


The fascinating thing about norms is that in any given situations, all parties involved are (somewhat) internally aware of the process and they can act accordingly. We can even deduce this (most probably) correct process in face of highly dynamic and nondeterministic situations with a multitude of variables. It seems like somehow we have converged very closely to a set of abstract, parameterized, context-based, and flexible scenario templates for interactions defined by norms that we use regularly as a coordination mechanism in the society. These scenario templates have placeholders for different roles each with their own script, and most of the time we can (almost) precisely interpret a situation, select a relevant and applicable template and instantly fill the parameters and role placeholders with qualified values and infer some sort of decision tree that tells us how an interaction can and should move forward. We even monitor and observe events and (own or others') actions and qualify them as traversal steps in these scenarios. Furthermore, with regards our part in a scenario, we seem to have internalized behavioral specifications or partial plans that allow us to enact and embody the role that we have filled with self.


Interpreting the context, selecting a relevant and applicable scenario template, filling in the variables, and qualifying observation are main components in utilizing scenario templates defined by norms. 
%Let us assume some examples
%, imagine again encountering a friend, but in two different contexts: once in their neighborhood where you typically see them every day and once somewhere on the other side of the world; while the first template perfectly matches the saying ``hi'' scenario context, the second one has a completely different context and will result in a substantially different concretization. 
Take a simplistic example 
%that is later on , when we encounter a friend, among all the actions available to us we choose to greet them and among all the words in the dictionary we choose saying ``hi'' with the highest probability, and, if all the conditions are right, we expect to hear ``hi'' back in a few seconds. Another example 
that is later on used in this thesis namely a sale transaction; regardless of the specific contexts, the norms for a typical sale transaction specify (almost globally accepted) predefined roles (buyer/seller), variables (item/price) and actions (offer/accept/pay/deliver) related to it. For an example of interpreting the context and selecting an applicable scenario take in-person shopping versus online shopping, while there are obvious differences in the variables and actions involved, we can think of them as the same thing that is a sale transaction with the same abstract template. For an example of filling in the variables, we can easily qualify actors and put them in proper roles, in fact the role of the seller in a sale transaction can be filled with ourselves, another person, a company or more interestingly even an automated software in the form of a webshop. The same level of abstraction also applies for qualifying observations and expectations, receiving cash from an in-person buyer can be qualified as the same action as a notification of a transfer from a banking system, and the expectation of receiving an item in the next few moments from an in-person seller is the same as the expectations of receiving an oversea delivery in the next few weeks.

The scenario templates defined by norms are also highly dynamic and flexible. During an interaction we are constantly monitoring the scenario for deviations from what we expect to occur. Some deviations can be negligible: while somewhat unexpected, they do not change the context of the interaction completely, like not receiving a delivery when it was supposed to arrive, maybe there was a delay. However, some deviations completely change the context, like never receiving an item you payed for and not being able to contact the seller. What is interesting about deviations is that we even have modular and predefined contingency plans for when a concrete scenario is not played according to expectations of its template. These plans can vary from attempting to fix the concrete scenario, like waiting a bit longer than expected for a delivery, to revising some beliefs that resulted in the unexpected situation like removing a particular actor from the qualifications that makes them suitable for the seller role to even rewriting the qualification or behavioral rules altogether like never buying anything from any unrecognized webshops.

\paragraph{Regulations and System Compliance}
Regulations are a form of norms that are (partially) concretized, written as a normative text and are enforced by some authoritative body.  Cyber-infrastructural systems that affect members of a society, should be compliant to the norms and regulations of that society. There are multiple ways to ensure compliance in a system, depending on the size, complexity of the system, norms that govern it and the governance bodies that are concerned with monitoring and regulating it. Some policies can be operationalized in terms of access control. For example some software systems with pre-defined and static internal policies can be developed with the policies already hard-coded in them, intuitively, this is most efficient approach whilst being the least maintainable. There are systems that use run-time policy-making protocols like LDAP to make sure the system is running in a compliant manner by e.g., taking permission from policy enforcement points that take advise from policy decision points which in turn retrieve policies from policy information point. Then by modifying the policies in information points, the system administrators can control enforced policies without changing the system itself. These approaches are arguably suitable for many systems, until the point that policies in question become more complicated than simple non-conflicting, non-contradicting, easy to interpret regulative norms defined in terms of deontic notions of permissions, obligations and prohibitions implemented in the form of ex-ante access rules. As it is observed in the literature, and multiple examples presented in this thesis show that social norms do indeed become more complex much sooner than anticipated by many works and in fact are vastly simplified in current compliance-checking approaches. 


Firstly, norms are more than a set of formal rules extracted from a legislative text: they emerge from multiple sources with different degrees of priority, they require interpretation to be encoded and qualification to be applied within a social context. Furthermore, they continuously adapt, in both expression and application~\cite{Boella2014APractice}. Furthermore, in any given context there are multiple normative sources may be concurrently relevant, and/or multiple interpretations of the same normative sources may be available (e.g. retrieved from previous cases), and these may be possibly conflicting. Also, not all regulations are ex-ante norms about controlling actions prior to the performance of the action, instead, they are ex-post rules about what oght to be if certain actions have been performed already. This makes conflict resolution and interpretation and concepts like consequences of actions detrimental parts of compliance-checking.


Furthermore, norms are traditionally distinguished between \textit{regulative} and \textit{constitutive} norms~\cite{Searle1969,Boella2004RegulativeSystems,Sileno2015}. Regulative norms regulate behaviors that exist independent of the norms and are generally prescribed in terms of \textit{permissions}, \textit{obligations} and, \textit{prohibitions} (e.g. traffic regulations). Constitutive norms determine that some entity (e.g. an object, a situation, a certain agent, a certain behaviour) ``counts as'' something else, creating a new institutional entity that does not exist independently of these norms. (for example, the concept of marriage or money as a legal means of payment). The concept of institutional power is particularly relevant in the context of constitutive norms, as it is used to ascribe institutional meaning to  performances (e.g. raising a hand counts as a bid during an auction). A conceptual framework that instead contains both deontic and potestative dimensions is the one proposed by Hohfeld \cite{hohfeld1917fundamental}, whereas deontic logic, although much more popular, by definition focuses on regulative norms. This work tries to take a step in the direction of taking these concepts into account in the process of designing compliant systems, or, modifying existing systems to be compliant.

\paragraph{Models of Agency}
Agency is another concept that is central to this work. Software systems are generally considered to be void of real agency and are controlled through specific control structures, even if those structures consist of probabilistic flows that can change as the system interacts with its environment like in case of reinforcement learning they still have a controll structure. This assumption becomes fuzzy when it comes to computational models of social actors, like executable models that should simulate human behavior, as without access to fully developed artificial general intelligence, it is hard to claim that a software agent has agency. Then the question is how do we model such actors? There are two main approaches used in the field to model concepts of agency such as objectives and preferences, namely these approaches are (1) modelling agents in terms of variables and mathematical functions, and, (2) modelling agents in terms of agent-oriented programs often in the form of logic-based behavioral specifications. 


The first approach has many advantages, firstly, mathematical models are often more efficient in execution resulting in them being more scalable. Also, defining agents in terms of a few variables makes them easier for experimentation particularly in simulations, and, it also makes them much easier to manipulate at run-time, to simulate adaptation and learning processes. In general, when the purpose of a research is to study the outcome of a social phenomena in certain settings without regards to the model of the input agent, mathematical models are suitable. An example of these approaches is the simulations of disaster response systems, where purpose of the study is to find the optimal settings or infrastructures to minimize damage. In these cases the agents are generally considered to have mathematical specifications that encompass a statistically realistic representation of how a person will respond to a natural disaster by avoiding danger or seeking safety as their utility function\cite{Something}.


Where mathematical models fall short and logic-based programmable models are advantageous is studies where expressivity and transparency of the models is the main point. In these cases, even if the execution is less efficient and adaptability is harder to achieve, it is still worth to have self-explanatory and readable models. These are the cases where the purpose of the study is to analyze the agents without much regard to optimizing an outcome and often there is no obvious utility to optimize in the setting anyways. The study of norms and normative agents is one domain where these approaches are far more adopted, as the transparency of the decision-making becomes more important in analyzing and explaining (non-)compliant behaviors that goes beyond utility functions. This research intrinsically falls into this category, and, as it will be presented in later chapters, the Belief-Desire-Intention (BDI) model of agency \cite{Something} is utilized in modelling agents. By specifying agents in terms of human-related attitudes, BDI agents are suitable in modelling social agents. More interestingly, majority of the concepts defined in the previous section about norms, like presence of behavioral specifications that can be matched and concretized in certain situations, or having contingency plans for failures are all already (partially) identified by BDI models to various degrees of success.


\section{Motivation and Research Question}
The overarching motivation of this research started has been in creation of a digital data market-places (DMP) specially in the field of logistics data. The need for data is exponentially increasing in all research and industries, and an environment like a DMP can provide different parties a place to share (or buy and sell) scientific or corporate data with each other. Although the presence of such environment can vastly improve the efficiency of data-oriented research and even prevent issues like data monopolies, there are certain risks involved, such as privacy issues, competitive corporate advantages, and, legal problems. As more governments (and other authoritative bodies) are implementing regulations to govern data transactions, such sharing environment needs to be compliant with these regulations, also actors in these markets like organizations and companies may have ad-hoc contractual agreements about data-sharing, and, they may also have internal policies like user agreements about how they can share data. The complexities of the market-places alongside the requirement for compliance results in a need for approaches to automate governance. This work started by focusing exactly that, how do we do make data-sharing compliant? 

However, during this research it turned out that it is not only data-sharing use-cases that are in need of such methods. More aspects of our day-to-day life are being \textit{controlled} by automated processes; visa applications, job applications, credit placements, mortgage and insurance are just a few examples and one can only imagine that the list will continue to grow as time goes on. Even if data-sharing regulations are a relatively new phenomena, when taken in a broader sense, there are already regulations implemented for other domains that are now becoming more automated, and when then the decisions are made in an automated manner, then the question becomes how do we make any automated process compliant to arbitrary (regulative and constitutive) normative sources? This thesis takes a subset of this question as its main purpose, by assuming modelling and model execution as a principal step in system design and policy making:

\begin{itemize}
    \item \textbf{Main Research Question:} How can we model a norm-governed socio-technical system?
\end{itemize}


\subsection{Approach and Scope of this Thesis}
The first step in defining the scope of this thesis is to break the main research question structurally by further defining what it is that needs to be modelled: a norm-governed socio-technical system. The main components are norms, social setting and software/infrastructure that create the system as a whole, 
%but also there is the assumption of openness, which means the system that is being modelled interacts with, affects and get affected by an external undefined environment. 
Then the main research sub-questions should intuitively become how to model each of these three components. But, unfortunalty it is not that simple as these concepts are not at the same level of abstraction for the purposes of this research. 
%Starting from the concept of social setting, it is arguably the main focus of this work, modelling social agents seems to be one of the keys to modelling such systems, but why? 
Take the concept of norms, while modelling them is essential for the whole picture, it is still social agents that act as the governance bodies that regulate other entities of the system, and in effect, it is social agents that are being governed. Indeed, it is very hard to talk about norm-related concepts like sanctions without referring to human-related concepts like \textit{intention} and it is rare to see a an infrastructure being punished for non-compliant actions without referring to a social agent (a person, company, institution, or organization) as the main liable entity for those actions. In effect, norms as are studied in this research do not have much meaning by themselves and without the social setting. The same logic goes for software and infrastructural components, while being another important part of the system model, it is still their usage by and effects on social agents that is being scrutinized. Furthermore, while software modelling is a mature and advanced field of study, the purposes of this work need a more realistice view of software systems. And such realistic view reveals that software and infrastructural systems are very technology-dependant subjects and high-level models of them are rarely complete without their low-level technological nuances. It is again hard to talk about a data transfer without mentioning if that data is personal, anonymized, plain or encypted, or the transmition channel is secure or not. In summary, in the context of this thesis, social agents are the glue that hold the models together. As this modelling is for the purpose of analysing the system as a whole, the expressivity and transparency of agent models is a requirement.

\begin{itemize}
    \item How to create expressive models of social agents?
\end{itemize}

The short answer to this question is agent-based programming is the most suitable approach to create models of agents. The alternate approach that is mathematical models while having many advantages are simply not transparent enought at the required level of abstraction for our use-cases: it is not easy to model intention for example in an effectively optimizer agent that has a utility function driving its decisions. The approach adopted apin this work to agent-based programming is the the Beleif-Desire-Intention model. While there are multiple BDI frameworks introduced in the literature, after much delibaration it turned out they did not meet the requirements of this study, maybe the most important one being modularity and interoperability. This meant that in the process of this thesis, a BDI framework called AgentScript Cross-Compiler (ASC2) was designed and developed with these requirements in mind. The ASC2 framework is introduced mostly from an engineering perspective in Chapter \ref{ch:agere}. 

Next, there is the issue of how to model software and infrastructural entities. This question effectively falls out of the scope of this thesis, instead, the focus is on their utilization by social agents. Important steps are taken to provide a high level of interoperability for social agents modelled in ASC2 to virtually any type of model of software and infrastructural entities that is being used by the designer and policy-maker, including the actual real entities.

\begin{itemize}
    \item How can social agents utilize software and infrastructural models?
\end{itemize}

The answer to this question is in the \textit{Cross-Compiler} part of ASC2. As it will be pointed out in Section \ref{xxx}, after compilation ASC2 agents are technically JVM programs meaning they are indeed interoperability with any system that for example a Java program can utilize or communicate with. 



Next is the notion of norms, modelling them including social norms, regulations, contractual agreements, internal policies is an important part of the modelling that this thesis aims at. However, while different approaches and ideas about modelling norms is explored in this work, defining any novel approach to do so falls out of the scope of this thesis for both theoretical and practical reasons. The practical reason is mainly that other threads of dedicated research were being performed on this matter within the research group that both affected and utilized this work. The theoretical reason is that there is no single best way to model norms, there are various levels of abstraction that norms can be modelled at for different use-cases and this work stays agnostic to this subject. However, the question that is present in this work is the concept of norms from the perspective of social agents:

\begin{itemize}
    \item How can social agents reason with and about norms?
\end{itemize}

This matter is referred to as study of \textit{normative} agents. The main concern of this thesis has been to study the interactions between social agent and norms, including compliant and non-compliant bahaviors, the relation between norm compliance and autonomoy, and, behavior in face of conflicting normative situation. This is addressed in section \ref{xxx} with introducing the concept of normative advisors, a flexible approach that can be utilized by social agents to have an understanding of normative positions in their environment without reducing their autonomoy.




\section{Policy-Making Scenarios}


\section{Research Collaborations}
Peter, Xin, Lu-Chi, Giovanni, Lu, Sarah, and, Thomas go here.